Software architectures to integrate workflow engines in science gateways

### 2016-05-15 paper.tex

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
other specific features. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down to
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores, as explained in the Table caption. For each
criterion, a global score is computed by summing up the individual
scores for each metric. We ensure that the different metrics used to
calculate the global score for a criterion are comparable, so that
they can be combined meaningfully. The criteria and metrics are
explained hereafter.

@@

Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

@@

\emph{Extensibility} measures the difficulty to replace or add
elements to the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down to 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
 \item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. 
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
 Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  potential for a single gateway instance to use more than one engine
  instance simultaneously. Workflow engines may require important
  amounts of resources when several workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load among several engines. \texttt{S$_1$}=0 means
  that
  adding a new engine instance is inherently supported in the
  architecture. In this case, elastic engines can be implemented
  through some kind of auto-scaling mechanism to control the number of
  engine instances in the architecture. \texttt{S$_1$}=1 means that
  multiple engine instances requires specific developments in the
  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
  engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

@@

Differences in other specific features should not be neglected. Task encapsulation, nested
workflows (with service invocation) and workflow conversion (with service
invocation) are the only architectures that intrinsically support meta-workflows. Besides,
tight integration is the only architecture that allows
fine-grained debugging, which may be
critical to efficient user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes since
it assumes that all criteria have equal weights while real systems
would favor some of them. Overall, task encapsulation performs a bit
better than the other architectures and nested workflows stand a bit
behind for the reasons mentioned previously.
\begin{table*}
\footnotesize
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Conversion} \\
  Integration (global score)    &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D299}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (global score) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (global score)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (global score) & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99DA99}\textbf{1.6}
                                    & \cellcolor[HTML]{99B899}\textbf{2.6}
                                    & \cellcolor[HTML]{99FF99}\textbf{0.5}
                                    & \cellcolor[HTML]{99BF99}\textbf{2.4}
                                    & \cellcolor[HTML]{999999}\textbf{3.5}
                                    & \cellcolor[HTML]{99DE99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized global scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained as in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

@@

A few limitations should be considered when using the results of our
evaluation. First, our evaluation methods aimed to provide comparative
metrics, however the high-level of abstraction used to derive these
metrics hinders the complexity of real systems to some
extent. In particular, the
presented architectures are abstract patterns that may be mixed
together in actual systems. The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hybrid architectures.

Moreover, all the interactions involved in the architectures were
treated equally, whereas some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation). However, without entering into the details of a particular
system, quantification of the robustness or the amount of development
associated with each interaction and software component will hardly be
precise.

@@

The abstract nested workflows and workflow conversion patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations.

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, many workflow engines are
already available as web services, which tends to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

There is abundant literature describing specific workflow engines and
systems; some of the main references are cited in
Section~\ref{sec:introduction-engines}. A few works described the
architecture and features of workflow system at an abstract level. For
instance, the early work~\cite{yuTaxonomy} proposes a generic
architecture for grid workflow systems that is based on the workflow
reference model defined by the Workflow Management
Coalition\footnote{\url{http://www.wfmc.org}}.  Deelman et
al. \cite{deelman2009workflows} have further characterized the
features of workflow systems. These two works, however, do not mention
science gateways.

@@

% Impact of the work
To the best of our knowledge, our work is the first to systematically
review and compare software architectures  to integrate workflow
engines into science gateways. So far, the literature on science gateways and
workflow engines has focused on the description of particular systems,
or on the presentation of a particular architecture.  Instead, our
analysis abstracts and evaluates architectural patterns independently
from any particular system. It provides general insight to integrate
science gateways and workflow engines, which we hope will be useful
for software architects.

>>>>

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
other specific features. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each
criterion, a global score is computed by summing up the individual
scores for each metric. We ensure that the different metrics used to
calculate the global score for a criterion are comparable, so that
they can be combined meaningfully. The criteria and metrics are
explained hereafter.

@@

Science gateway \emph{integration complexity} is measured through the number of software components and their interactions. It breaks
down into the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not developed nor maintained by the groups who
integrate workflow engines in science gateway. We support its inclusion here because integrating an
infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), therefore increasing complexity. The two metrics are summed to obtain a
global score that measures the
total number of software pieces affected by integration.

@@

\emph{Extensibility} measures the difficulty to replace or add
elements to the architecture. It is determined by the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down into 4 metrics depending on the type of element (engine, workflow or infrastructure) that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine (\texttt{E$_1$}): number of interactions or
  components that need to be modified to integrate a new type of
  workflow engine in the architecture. Workflow engines belong to
  different types when they cannot be invoked using the same
  implementation of the interactions.
 \item Engine version upgrade (\texttt{E$_2$}): number of interactions
  or components modified to integrate a new version of a workflow
  engine in the architecture, assuming that another version of the
  same engine type is already available. Different versions of a
  workflow engine can be invoked using the same interaction
  implementation(s). When this is not the case, the different versions
  are considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc. In practice, E$_2$=0 when the workflow
  engine is wrapped in a service or in an agent, and E$_2$=1 when it is
  wrapped in another component.  
\item New workflow (\texttt{E$_3$}): number of interactions required
  to integrate a new workflow in the architecture. Adding a new
  workflow is a very common operation that does not require modifying
  software components or interactions, assuming that the engine type
  and version required to execute this workflow are already
  available. In most cases, adding a new workflow requires only
  interactions \texttt{a$_1$} and \texttt{a$_2$}, but \texttt{g}
  is required as well when workflow conversion is used. 
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components affected for the integration of a new type of infrastructure in
  the architecture. Adding a new infrastructure typically aims at providing more
  computing or storage power, enabling access to specific types of resources
  (e.g., GPUs, clouds), or  enforcing execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The four metrics are summed to obtain a global score that measures the
difficulty to extend the architecture. Note that some extensions may
affect several metrics in practice. For instance, adding a new type
of engine may help integrating new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and the infrastructure.

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
 Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  potential for a single gateway instance to use more than one engine
  instance simultaneously. Workflow engines may require important
  amounts of resources when several workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load among several engines. \texttt{S$_1$}=0 means
  that
  adding a new engine instance is inherently supported in the
  architecture. In this case, elastic engines can be implemented
  through some kind of auto-scaling mechanism to control the number of
  engine instances in the architecture. \texttt{S$_1$}=1 means that
  supporting multiple engine instances requires specific developments in the
  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
  engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to implement task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

@@

Differences in other specific features should not be neglected. Task encapsulation, nested
workflows (with service invocation) and workflow conversion (with service
invocation) are the only architectures that intrinsically support meta-workflows. Besides,
tight integration is the only architecture that allows
fine-grained debugging, which may be
critical for efficient user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes, since
it assumes that all criteria have equal weights while real systems
would favor some of them. Overall, task encapsulation performs a bit
better than the other architectures and nested workflows stand a bit
behind for the reasons mentioned previously.
\begin{table*}
\footnotesize
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Conversion} \\
  Integration (global score)    &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D299}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (global score) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (global score)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (global score) & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99DA99}\textbf{1.6}
                                    & \cellcolor[HTML]{99B899}\textbf{2.6}
                                    & \cellcolor[HTML]{99FF99}\textbf{0.5}
                                    & \cellcolor[HTML]{99BF99}\textbf{2.4}
                                    & \cellcolor[HTML]{999999}\textbf{3.5}
                                    & \cellcolor[HTML]{99DE99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized global scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained as in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

@@

A few limitations should be considered when using the results of our
evaluation. First, our evaluation methods aimed to provide comparative
metrics, however the high-level of abstraction used to derive these
metrics hinders the complexity of real systems to some
extent. In particular, the
presented architectures are abstract patterns that may be mixed
together in actual systems. The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. Nevertheless, the criteria discussed in this
work would still be applicable for broadly comparing and categorizing such cases of
hybrid architectures.

Note also that all the interactions involved in the architectures were
treated equally, whereas some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation). However, without entering into the details of a particular
system, quantification of the robustness or the amount of development
associated with each interaction and software component will hardly be
precise.

@@

The abstract nested workflows and workflow conversion patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures have already been
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations.

The particular technical or historical context of a science gateway
project may also influence the choice of an architecture to
integrate workflow engines. For instance, many workflow engines are
already available as web services, which tends to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

There is abundant literature describing specific workflow engines and
systems; some of the main references are cited in
Section~\ref{sec:introduction-engines}. A few works described the
architecture and features of workflow systems at an abstract level. For
instance, the early work~\cite{yuTaxonomy} proposes a generic
architecture for grid workflow systems that is based on the workflow
reference model defined by the Workflow Management
Coalition\footnote{\url{http://www.wfmc.org}}.  Deelman et
al. \cite{deelman2009workflows} have further characterized the
features of workflow systems. These two works, however, do not mention
science gateways.

@@

% Impact of the work
To the best of our knowledge, our work is the first to systematically
review and compare software architectures  to integrate workflow
engines into science gateways. So far, the literature on science gateways and
workflow engines has focused on the description of particular systems,
or on the presentation of a particular architecture.  Instead, our
analysis abstracts and evaluates architectural patterns independently
from any particular system, providing general insight about ways to integrate
science gateways and workflow engines. These insights will be valuable
for software architects when considering alternatives for the architecture and evaluating the impact of their decisions.
